The watershed transform in ITK - discussion and new developments

### 2006-05-15 article/Article.tex
more of the article
darcs-hash:20060515142026-2fc9d-fa340f3526b977f262c2cfdfd5185152b04cd742.gz

Another major advance made soon after by Meyer and Beucher
\cite{Beucher93a,Meyer1994a}. Meyer defined the watershed transform in terms of {\em
topographic distances} through {\em lower complete} images. These
definitions rigorously defined the behavior in the presence of
plateaus and lead to some new algorithms based on graph theory. Two
examples of these algorithms are {\em hill-climbing} and {\em image
integration} and will be discussed in more detail in this report.
Figures \ref{fig:cthead1-markers}, \ref{fig:circle}and \ref{fig:difficult-cases} illustrate
the possible problems with plateaus.

@@

\subsection{The topological watershed}
An alternative approach, known as the topological watershed, has
recently been developed \cite{}. The topological watershed modifies
the topology of the surface so that all pixels that are not watershed
lines become regional minima and watershed lines are modified so that
they are more suitable for use in measuring saliency of boundaries
between regions. Algorithms implementing the topological watershed
also exploit priority queues, but don't appear to be able to avoid the
need for explicit lower completion or reconstruction by erosion.

@@

The ITK
watershed implementation provides a way of addressing this problem
using a measure of significance of the boundary and allowing a
threshold to be applied first to reduce the number of local minima.

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=3.2]{level}}
\subfigure[No minima level]{\includegraphics[scale=3.2]{level00-rgb}}
\subfigure[Minima level = 10]{\includegraphics[scale=3.2]{level10-rgb}}
\subfigure[Minima level = 20]{\includegraphics[scale=3.2]{level20-rgb}}
\subfigure[Minima level = 30 and Minima level = 40]{\includegraphics[scale=3.2]{level30-rgb}}
\subfigure[Minima level = 50]{\includegraphics[scale=3.2]{level50-rgb}}
\subfigure[markers]{\includegraphics[scale=3.2]{level-markers-rgb}}
\subfigure[Watershed from markers]{\includegraphics[scale=3.2]{level-markersMf-rgb}}
\caption{Different results of morphological watershed on an articial image. The labeled image are overlayed to the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image. All the watersheds were performed with $MarkWatershed = true$ and $FullyConnected = false$.}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=3.2]{level}}
\subfigure[No minima level]{\includegraphics[scale=3.2]{level00-itk-rgb}}
\subfigure[Level = 0.1]{\includegraphics[scale=3.2]{level10-itk-rgb}}
% \subfigure[Level = 0.2]{\includegraphics[scale=3.2]{level20-itk-rgb}}
\subfigure[Level = 0.4]{\includegraphics[scale=3.2]{level40-itk-rgb}}
\subfigure[Level = 0.63]{\includegraphics[scale=3.2]{level63-itk-rgb}}
\subfigure[Level = 0.65]{\includegraphics[scale=3.2]{level65-itk-rgb}}
\subfigure[markers]{\includegraphics[scale=3.2]{level-markers-rgb}}
\subfigure[Watershed from markers]{\includegraphics[scale=3.2]{level-markers-itk-rgb}}
\caption{Different results of ITK's watershed on an articial image. The labeled image are overlayed to the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image.}
\end{center}
\end{figure}

@@

\subsection{Watershed from selected minima}
An alternative approach leads us to the concept of markers mentioned
earlier. We will also see that the thresholding approach used by ITK
may be considered as a specific case of the same approach.

@@

Contrary to the previous algorithm, the label in the output image
can't be used to determine if the pixel has already been added to
the queue, and so we need to maintain this status by another way.
It is the role of $S$ in Figure \ref{MeyerAlgorithm}. We'll discuss
in \ref{sect:implementation} how this status can be implemented.

@@

We provide with this article two new classes:

@@

\section{Timing and performance}
The relative performance of the new filters are shown in
Table~\ref{tbl:perf}. The timings were obtained using the {\tt
./perf3D images/ESCells.img}. The times for steps involved in a
morphological watershed are shown in the first few columns, with the
time from a combined filter and the itk watershed shown last. It
should be noted that the itk watershed does not allow connectivity to
be modified by the user - it always uses face connected mode and
doesn't have an inbuilt mechanism for boundary selection. The existing
ITK watershed is therefore somewhat slower than the new implementation
when used to compute a traditional watershed. The new implementations
are not easily threadable and therefore may prove to be slower on some
systems. However, as discussed in earlier sections, the new
implementation produces more stable results than the rainfall model
and can be used in a marker based approach.

@@

@@

\subsection{Pronucleus segmentation in bovine embryo}

In this example, we use the watershed filter to split two nearly round
objects: the male and female pronucleus in a bovine embryo, around 22
hours after fecondation.

@@

\small \begin{verbatim}
ws = itk.MorphologicalWatershedImageFilter.IUC2IUC2.New(idistance,
                                                        MarkWatershed=False,
                                                        Level=10)
mask = itk.MaskImageFilter.IUC2IUC2IUC2.New(ws, fill)
overlay = itk.LabelOverlayImageFilter.IUC2IUC2IRGBUC2.New(reader, mask, Opacity=0.5)
\end{verbatim} \normalsize

@@

\small \begin{verbatim}
ws2 = itk.WatershedImageFilter.IUC2.New(idistance, Level=0.5)
relabel = itk.RelabelComponentImageFilter.IUL2IUC2.New(ws2)
mask2 = itk.MaskImageFilter.IUC2IUC2IUC2.New(relabel, fill)
overlay2 = itk.LabelOverlayImageFilter.IUC2IUC2IRGBUC2.New(reader, mask2, Opacity=0.5)
\end{verbatim} \normalsize

@@

>>>>

Another major advance made soon after by Meyer and Beucher
\cite{Beucher93a,Meyer1994a}. Meyer defined the watershed transform in terms of {\em
topographic distances} through {\em lower complete} images. These
definitions rigorously defined the behavior in the presence of
plateaus and lead to some new algorithms based on graph theory. Two
examples of these algorithms are {\em hill-climbing} and {\em image
integration} and will be discussed in more detail in this report.
Figures \ref{fig:cthead1-markers}, \ref{fig:circle} and \ref{fig:difficult-cases} illustrate
the possible problems with plateaus.

@@

\subsection{The topological watershed}
An alternative approach, known as the topological watershed, has
recently been developed \cite{topologicalWatershed}. The topological watershed modifies
the topology of the surface so that all pixels that are not watershed
lines become regional minima and watershed lines are modified so that
they are more suitable for use in measuring saliency of boundaries
between regions. Algorithms implementing the topological watershed
also exploit priority queues, but don't appear to be able to avoid the
need for explicit lower completion or reconstruction by erosion.

@@

A common way to address this problem is to use a ignore the minima
with a too small dynamic. Figures \ref{fig:artificial-itk} and
\ref{fig:artificial-morpho} illustrate the effect of this technique.
An usage example is shown in Section \ref{sect:pronucleus}.
The ITK watershed implementation also allow a threshold to be applied
first to reduce the number of local minima.

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=3.2]{level}}
\subfigure[No minima level]{\includegraphics[scale=3.2]{level00-rgb}}
\subfigure[Minima level = 10]{\includegraphics[scale=3.2]{level10-rgb}}
\subfigure[Minima level = 20]{\includegraphics[scale=3.2]{level20-rgb}}
\subfigure[Minima level = 30 and Minima level = 40]{\includegraphics[scale=3.2]{level30-rgb}}
\subfigure[Minima level = 50]{\includegraphics[scale=3.2]{level50-rgb}}
\subfigure[markers]{\includegraphics[scale=3.2]{level-markers-rgb}}
\subfigure[Watershed from markers]{\includegraphics[scale=3.2]{level-markersMf-rgb}}
\caption{Different results of morphological watershed on an articial image. The labeled image are overlayed to the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image. All the watersheds were performed with $MarkWatershed = true$ and $FullyConnected = false$.\label{fig:artificial-itk}}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=3.2]{level}}
\subfigure[No minima level]{\includegraphics[scale=3.2]{level00-itk-rgb}}
\subfigure[Level = 0.1]{\includegraphics[scale=3.2]{level10-itk-rgb}}
% \subfigure[Level = 0.2]{\includegraphics[scale=3.2]{level20-itk-rgb}}
\subfigure[Level = 0.4]{\includegraphics[scale=3.2]{level40-itk-rgb}}
\subfigure[Level = 0.63]{\includegraphics[scale=3.2]{level63-itk-rgb}}
\subfigure[Level = 0.65]{\includegraphics[scale=3.2]{level65-itk-rgb}}
\subfigure[markers]{\includegraphics[scale=3.2]{level-markers-rgb}}
\subfigure[Watershed from markers]{\includegraphics[scale=3.2]{level-markers-itk-rgb}}
\caption{Different results of ITK's watershed on an articial image. The labeled image are overlayed to the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image.\label{fig:artificial-morpho}}
\end{center}
\end{figure}

@@

\subsection{Watershed from selected minima}
\label{sect:ws-selected-minima}
An alternative approach leads us to the concept of markers mentioned
earlier. We will also see that the thresholding approach used by ITK
may be considered as a specific case of the same approach.

@@

Contrary to the previous algorithm, the label in the output image
can't be used to determine if the pixel has already been added to
the queue, and so we need to maintain this status by another way.
It is the role of $S$ in Figure \ref{MeyerAlgorithm}. We'll discuss
in Section \ref{sect:implementation} how this status can be implemented.

@@

  \subsection{With ITK watershed}

The watershed from markers can be implemented with the ITK watershed,
by using the technique of {\em minima imposition} exposed in Section
\ref{sect:ws-selected-minima}. Several steps are required:
\begin{itemize}
  \item Impose the minima to the input image, with the
\verb$MinimaImpositionImageFilter$
  \item Perform the watershed with the \verb$WatershedImageFilter$
  \item Restore the labels produced by the \verb$WatershedImageFilter$.
\verb$WatershedImageFilter$ produce arbitrary labels which doesn't match
the labels in the marker image. There is currently no filter able to
find the label correspondance in ITK, so this step must be performed by
hand with a code like the one before. We can use a
\verb$ChangeLabelImageFilter$ to change the labels.
\small \begin{verbatim}
filter->Update();

typedef itk::ChangeLabelImageFilter< LIType, IType > ChangeLabelType;
ChangeLabelType::Pointer changeLabel = ChangeLabelType::New();
changeLabel->SetInput( filter->GetOutput() );

itk::ImageRegionConstIterator< LIType > wit( filter->GetOutput(),
                filter->GetOutput()->GetLargestPossibleRegion() );
itk::ImageRegionConstIterator< IType > mit( reader2->GetOutput(),
                filter->GetOutput()->GetLargestPossibleRegion() );
for ( mit.GoToBegin(), wit.GoToBegin();
  !mit.IsAtEnd();
  ++wit, ++mit )
  {
  if( mit.Get() != 0 )
    { changeLabel->SetChange( wit.Get(), mit.Get() ); }
  }
\end{verbatim} \normalsize
\end{itemize}

This approach has several drawbacks:
\begin{itemize}
  \item it doesn't work if there some markers not separated by background
pixels - they are considered as a single marker
  \item restoring the labels is not included in a filter and so break the
ITK pipeline model
  \item it's less efficient, as shown in Section \ref{sect:perf}
  \item it gives poor results with plateaus
\end{itemize}

A full example of implementation can be found in \verb$iwsm.cxx$.

@@

\section{Timing and performance}
\label{sect:perf}

  \subsection{Classical watershed}
\label{sect:perf-classical}

The relative performance of the new filters are shown in
Table~\ref{tbl:perf}. The timings were obtained using the {\tt
./perf3D images/ESCells.img}. The times for steps involved in a
morphological watershed are shown in the first few columns, with the
time from a combined filter and the itk watershed shown last. It
should be noted that the itk watershed does not allow connectivity to
be modified by the user - it always uses face connected mode and
doesn't have an inbuilt mechanism for boundary selection. The existing
ITK watershed is therefore somewhat slower than the new implementation
when used to compute a traditional watershed. The new implementations
are not easily threadable and therefore may prove to be slower on some
systems. Moreover, this results doesn't show the performance in the
case where the user modify the \verb$Level$ parameters of the ITK
filter, case where the time needed to update the filter should be a lot
shorter. However, as discussed in earlier sections, the new
implementation produces more stable results than the rainfall model
and is more dedicated to a marker based approach.

@@

  \subsection{Marker based watershed}
\label{sect:perf-marker}

The performance in case of a marker based approach are shown in Table
\ref{tbl:perf-marker}, and were obtained on the same system than the
ones in Section \ref{sect:perf-classical}, with the command {\tt 
./mperf3D images/ESCells.img images/ESCells-markers.tif}.
The ITK watershed use the implementation presented in Section
\ref{sect:implementation}. The extra steps of this implementation are
also shown in this table.

\begin{table}[htbp]
\centering
\begin{tabular}{cccccccc}
\hline
FC & MB & impose min &  itk watershed & match labels & change labels & total & Morph watershed \\
\hline
\hline
\end{tabular}
\caption{Execution times with marker based approach. Note the FC and MB refer to FullyConnected and MarkBoundary.\label{tbl:perf-marker}}
\end{table}

@@

\subsection{Pronucleus segmentation in bovine embryo}
\label{sect:pronucleus}
In this example, we use the watershed filter to split two nearly round
objects: the male and female pronucleus in a bovine embryo, around 22
hours after fecondation.

@@

\small \begin{verbatim}
ws = itk.MorphologicalWatershedImageFilter.IUC2IUC2.New(idistance,
                                                        MarkWatershed=False,
                                                        Level=10)
mask = itk.MaskImageFilter.IUC2IUC2IUC2.New(ws, fill)
overlay = itk.LabelOverlayImageFilter.IUC2IUC2IRGBUC2.New(reader, mask)
\end{verbatim} \normalsize

@@

\small \begin{verbatim}
ws2 = itk.WatershedImageFilter.IUC2.New(idistance, Level=0.5)
relabel = itk.RelabelComponentImageFilter.IUL2IUC2.New(ws2)
mask2 = itk.MaskImageFilter.IUC2IUC2IUC2.New(relabel, fill)
overlay2 = itk.LabelOverlayImageFilter.IUC2IUC2IRGBUC2.New(reader, mask2)
\end{verbatim} \normalsize

@@

\section{Related work}
This work have lead us to develop some other filters to:
\begin{itemize}
  \item Find regional extrema \cite{regionalExtrema}
  \item Surimpose a colored label image onto an image \cite{labelOverlay}
  \item Impose the minima \cite{minimaImposition}
  \item Label the connected component \cite{connectedComponent}
  \item Invert the intensity of an image \cite{invert}
  \item Enhance the morphological reconstruction \cite{recons}
  \item Compute some informations about the shape of the labeled objects (not published yet)
\end{itemize}
